---
title: "Data Viz 2 - Homework 1"
author: "Cosmin Catalin Ticu"
date: "3/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

Load the nycflights13 package and check what kind of datasets exist in the package, then create a copy of flights dataset into a data.table object, called flight_data.

```{r, message=FALSE, warning=FALSE}
library(data.table)
library(nycflights13)
library(lubridate)
library(ggplot2)
# for the last question
library(cowplot)
library(rnaturalearth)
library(googleway)
library(rgeos)
library(rnaturalearthdata)
library(ggspatial)
library(sf)
library(ggrepel)

flight_data <- data.table(nycflights13::flights)

str(flight_data)
```

Which destination had the lowest avg arrival delay from LGA with minimum 100 flight to that destination?

```{r}
temp <- flight_data[origin == 'LGA', list(number_of_flights = .N, 
                                  mean_arrival_delay = mean(arr_delay, na.rm = T)), 
            by = dest][order(mean_arrival_delay)]
temp[number_of_flights > 100,][1]
```

Which destination's flights were the most on time (avg arrival delay closest to zero) from LGA with minimum 100 flight to that destination?

```{r}
q2 <- temp[number_of_flights > 100, 
           list(dest, mean_arrival_delay, number_of_flights, 
                gicu = abs(mean_arrival_delay))][order(gicu)]
q2[, list(dest, number_of_flights, mean_arrival_delay)][1]

# correct because average arrival delay is actually negative 
# yet still closest to zero in absolute value (as seen in the gicu column above)
```

Who is the manufacturer of the plane, which flights the most to CHS destination?

```{r}
q3_temp <- flight_data[dest == 'CHS', list(number_flights = .N), 
                       by = list(tailnum, dest)][order(-number_flights)]

flight_data_planes <- data.table(nycflights13::planes)

merge(q3_temp, 
      flight_data_planes[, list(tailnum, 
                                manufacturer)])[order(-number_flights)][1]
```

Which airline (carrier) flow the most by distance?

```{r}
q4_temp <- data.table(nycflights13::airlines)

merge(flight_data[, list(total_distance_flown = sum(distance)), 
            by = carrier][order(-total_distance_flown)][1], q4_temp)

```

Plot the monthly number of flights with 20+ mins arrival delay!

```{r}
q5_temp <- flight_data[arr_delay > 20, list(date = make_date(year, month))]
ggplot(q5_temp, aes(x = date)) + geom_bar()
```

Plot the departure delay of flights going to IAH and the related day's wind speed on a scaterplot! Is there any association between the two variables? Try adding a linear model.

```{r,message=FALSE, warning=FALSE}
flights_weather <- data.table(nycflights13::weather)

q6_temp <- flight_data[dest == 'IAH', 
                       list(dep_delay, origin, dest, year, month, day)]

agg_weather <- flights_weather[, list(mean_wind_speed = mean(wind_speed)), 
                               by = list(origin, year, month, day)]

q6_answer <- merge(q6_temp, agg_weather)

ggplot(q6_answer, aes(dep_delay, mean_wind_speed)) + 
  geom_point() + 
  geom_smooth(method = 'lm')
```

Plot the airports as per their geolocation on a world map, by mapping the number flights going to that destination to the size of the symbol!

```{r}
map <- ne_countries(returnclass = "sf")

flight_airports <- data.table(nycflights13::airports)
geo_flight_data <- merge(flight_data[, list(number_of_flights = .N), by = dest], 
                         flight_airports[, list(faa, lat, lon)], 
                         by.x = "dest", by.y = "faa")

# To answer this question, the following resource was used
# https://www.r-spatial.org/r/2018/10/25/ggplot2-sf.html

ggplot(map) + 
  geom_sf() + 
  geom_point(data = geo_flight_data, aes(lon,lat, size = number_of_flights))
```

Misi's alternative answer

```{r, message=FALSE, warning=FALSE}
worldmap <- map_data('world')
ggplot() +
  geom_polygon(
    data = worldmap, 
    aes(x = long, y = lat, group = group), 
    fill = 'gray', color = 'black') +
  geom_point(
    data = merge(
      airports, 
      flight_data[, .(flights = .N), by = .(faa = dest)], 
      by = 'faa'), 
    aes(lon, lat, size = flights)) +
  theme_bw() +
  theme(legend.position = 'top') +
  xlab('') + ylab('') + 
  coord_fixed(1.3)

```